\twocolumn
\chapter*{Index of Symbols}
\addcontentsline{toc}{chapter}{Symbols}

% !TEX root = ./Almagest.tex
{\small
\begin{description}
\item $A$: azimuth.
\item $a$: major radius, altitude.
\item $a_S$: major radius of Sun.
\item $\alpha$: right ascension.
\item $\bar{\alpha}$: right ascension of mean sun. 
\item $b$: minor radius.
\item $\beta$: ecliptic latitude. 
\item $D$: lunar-solar elongation.
\item$\bar {D}$: mean lunar-solar elongation.
\item $\tilde{D}$: semi-mean lunar-solar elongation.
\item $\delta$: declination. 
\item $\delta_M$: parallax of Moon.
\item $e$: eccentricity.
\item $e_M$: eccentricity of Moon.
\item $e_S$: eccentricity of Sun.
\item $\epsilon$: inclination of ecliptic.
\item $E$: elliptic anomaly.
\item $\zeta$: radial anomaly. 
\item $\theta$: equation of epicycle.
\item $F$: argument of latitude.
\item $\bar{F}$: mean argument of latitude.
\item $\bar{F}_0$: mean argument of latitude at epoch.
\item $\bar{F}_M$: mean argument of latitude of Moon.
\item $\lambda$: ecliptic longitude. 
\item $\lambda_S$: ecliptic longitude of Sun.
\item $\bar{\lambda}$: mean longitude.
\item $\bar{\lambda}_0$: mean longitude at epoch. 
\item $\bar{\lambda}_M$: mean longitude of Moon.
\item $\bar{\lambda}_S$: mean longitude of Sun.
\item $L$: terrestrial latitude. 
\item $M$: mean anomaly.
\item $M_0$: mean anomaly at epoch.
\item $M_M$: mean anomaly of Moon.
\item $M_S$: mean anomaly of Sun.
\item $\mu$: epicyclic anomaly, parallactic angle.
\item $n$: rate of motion in mean longitude.
\item $\tilde{n}$: rate of motion in mean anomaly.
\item $\breve{n}$: rate of motion in mean argument of latitude.
\item $q$: equation of center.
\item $q_i$: lunar anomalies.
\item $R_E$: radius of Earth.
\item $R_M$: radius of Moon.
\item $R_S$: radius of Sun.
\item $r$: radial distance.
\item $\rho_S$: angular radius of Sun.
\item $\rho_M$: angular radius of Moon.
\item $\rho_U$: angular radius of Earth's umbra.
\item $t$: time.
\item $t_0$: epoch.
\item $T$: true anomaly.
\item $\tau$: orbital period.
\item$\varpi$: longitude of  perigee.
\item $\Omega$: longitude of ascending node.
\end{description}
\onecolumn
}